\chapter*{Appendices}
\markboth{Appendices}{}
\addcontentsline{toc}{chapter}{Appendices}
This appendix includes supplementary materials that support the main content of the report. It provides additional details, data, and resources that are relevant for a comprehensive understanding of the project.


\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}

\section{Source Code / GitHub Link}
\begin{itemize}
    \item Project Repository: https://github.com/Abdullah-Al-Ikram/Load-Shed-Forecast-Using-Hybrid-LSTM-GRU
\end{itemize}

This repository contains the complete source code for the load shedding forecasting project, including data preprocessing scripts, model training code, and evaluation metrics.

\section{Team Responsibilities}
\begin{itemize}
    \item Md. Abdullah Al Ikram:
    \begin{itemize}
        \item Data Collection and Preprocessing
        \item Model Development and Training
        \item Performance Evaluation and Analysis
        \item Algorithm Design and Implementation
        \item Hyperparameter Tuning and Optimization
        \item Result Visualization and Presentation
        \item Documentation and Report Writing
    \end{itemize}
    \item Tonoya Akhter Adhora:
    \begin{itemize}
        \item Literature Review and Background Research
        \item Documentation and Report Writing
    \end{itemize}

    \item Abdul Rizvan Alvi:
    \begin{itemize}
        \item Planning and Budget Estimation
        \item Analyze Standards and Design Constraints
    \end{itemize}
\end{itemize}

\section{Ethics and Compliance Statements}
\begin{itemize}
    \item Data Privacy: All data used in this project were obtained from publicly available sources or with proper authorization. No personally identifiable information (PII) was used or disclosed.
    \item Ethical Use of AI: The development and deployment of the Hybrid LSTM-GRU model were conducted in accordance with ethical guidelines for AI, ensuring fairness, transparency, and accountability.
    \item Compliance with Institutional Policies: The project adhered to the University of Asia Pacific's policies on research ethics and integrity.
    \item Environmental Considerations: Efforts were made to minimize the environmental impact of computational resources used during model training and evaluation.
\end{itemize}

\section{Evidences of CEP}
The project involved several Complex Engineering Problem (CEP) outlines how the project meets the criteria for addressing a Complex Engineering Problem (CEP) as defined by the accreditation standards. The following points summarize the evidences provided in the report:
\begin{itemize}
    \item Complex Engineering Problem (CEP) Addressed: The project addresses the complex engineering problem of accurately forecasting load shedding events in power grids, which involves handling large datasets, capturing non-linear temporal dependencies, and integrating multiple data sources.
    \item Methodological Framework: The use of a Hybrid LSTM-GRU model demonstrates the application of advanced deep learning techniques to solve a real-world engineering problem.
    \item Performance Metrics: The project includes comprehensive evaluation metrics (accuracy, precision, recall, F1-score, MAE, RMSE) to assess the effectiveness of the proposed solution.
    \item Comparative Analysis: A detailed comparison with baseline models (Pure GRU and Pure LSTM) highlights the advantages of the Hybrid approach in addressing the CEP.
\end{itemize}
\section{Evidences of CEA}
The project involved several Complex Engineering Activities (CEA) as outlined in Chapter \ref{cha:AccreditationComponents}. The following points summarize the key activities undertaken:
\begin{itemize}
    \item Data Collection and Preprocessing: Gathering and preparing historical load shedding data, weather data, and electricity demand data for model training.
    \item Model Development: Designing and implementing the Hybrid LSTM-GRU architecture for load shedding forecasting.
    \item Hyperparameter Tuning: Systematically optimizing model parameters to enhance performance.
    \item Performance Evaluation: Conducting rigorous testing and validation of the model using various metrics.
    \item Result Analysis: Interpreting the results and comparing them with existing approaches to demonstrate the effectiveness of the proposed solution.
\end{itemize}
