\vspace{-0.8cm}

% \chapter{Project Planning and Budget}
% \label{ch:planning}

This chapter presents an overview of the temporal planning and budget formulation processes undertaken for the project. It highlights how work activities were arranged and resource planning was conducted to maintain efficiency throughout the development phase.

\section{Project Timeline}
This section outlines the planned timeline for the project, detailing key phases, milestones, and deliverables. A Gantt chart is provided to visually represent the schedule and progression of tasks over the project duration.

\begin{table}[H]
\tiny
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|p{0.2\textwidth}|c|c|c|c|c|c|c|c|c|c|c|}
\hline
\rowcolor{gray!30}
\textbf{Project Phase} & \textbf{Jan} & \textbf{Feb} & \textbf{Mar} & \textbf{Apr} & \textbf{May} & \textbf{Jun} & \textbf{Jul} & \textbf{Aug} & \textbf{Sep} & \textbf{Oct} & \textbf{Nov} \\
\hline
Requirement Analysis & \cellcolor{gray!40} & \cellcolor{gray!40} &  &  & &  &  &  &  &  & \\
\hline
Data Collection \& Preprocessing &  & \cellcolor{gray!40} & \cellcolor{gray!40} & \cellcolor{gray!40} &  &  &  &  &  &  & \\
\hline
Model Development &  &  &  & \cellcolor{gray!40} & \cellcolor{gray!40}&  &  &  &  &  & \\
\hline
Model Training \& Optimization &  &  &  &  &\cellcolor{gray!40} & \cellcolor{gray!40} & \cellcolor{gray!40} &  &  &  & \\
\hline
Testing \& Validation &  &  &  &  & &  &  \cellcolor{gray!40} & \cellcolor{gray!40} & \cellcolor{gray!40} &  & \\
\hline
Results Analysis &  &  &  &  & &  &  & \cellcolor{gray!40} & \cellcolor{gray!40} & \cellcolor{gray!40} &  \\
\hline
Documentation \& Final Report &  &  &  &  & &  &  &  &  & \cellcolor{gray!40} & \cellcolor{gray!40} \\
\hline
\end{tabular}
\caption{Gantt Chart Showing Timeline of Major Phases}
\label{tab:gantt_chart}
\end{table}


\section{Budget Estimation}
This section provides a detailed budget estimation for the project, outlining anticipated costs associated with various resources and activities. The budget is structured to ensure transparency and effective resource allocation.

\begin{table}[tbh]
\centering
\resizebox{0.75\textwidth}{!}{
\begin{tabular}{lcrr}
\hline
\rowcolor{gray!30}
\textbf{Item/Resource} & \textbf{Qty} & \textbf{Unit Cost (BDT)} & \textbf{Total Cost} \\
\hline
Hardware Components & 1 & 50,000 & 50,000 \\
Software Tools (Google Colab) & 1 & 0 & 0 \\
Cloud Storage (Dataset) & 1 & 0 & 0 \\
Miscellaneous (optional) & 1 & 5,000 & 5,000 \\
\hline
\textbf{Total Estimated Cost} & & & 55,000 \\
\hline
\end{tabular}
}
\caption{Estimated Project Budget}
\label{tab:project_budget}
\end{table}


\subsection*{Guidelines}
All relevant cost categories shown in Table \ref{tab:project_budget}: 
\begin{itemize}
    \item Hardware/Equipment: PC, GPU, etc. This may be a one-time purchase or rental.
    \item Software/Services: Operating systems, Python, VS Code, Pycharm, MATLAB, Mathematica, Anaconda. This are mostly free of cost.
    \item Cloud Services: Cloud services (e.g., Google Colab Pro, AWS), specialized software licenses. This may be subscription-based.
    \item Miscellaneous: Internet costs, data acquisition fees, etc. This may include any other costs incurred during the project.
\end{itemize}


\section{Summary} 
This chapter has outlined the project timeline and budget estimation, providing a clear framework for managing time and resources effectively throughout the project lifecycle. The Gantt chart in Table \ref{tab:gantt_chart} illustrates the planned phases and milestones, while the budget breakdown in Table \ref{tab:project_budget} ensures transparency in resource allocation.